Efficient implementation of penalized regression\\for genetic risk prediction

### 2019-02-12 paper/PRS.tex
finished

\abstract{
Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  
The ``Clumping+Thresholding'' (C+T) approach is the most common method to derive PRS. C+T uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use. 
However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

{\color{red}
Applying penalized linear regression to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalability and strong predictive power, even for highly polygenic traits.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
This demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available.

In conclusion, our study demonstrates that penalized regression can provide discriminative polygenic risk scores, while being applicable to large-scale individual-level data because of an highly efficient implementation available in R package bigstatsr.
}
}

@@

Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that, for a 10\% disease prevalence, the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
}

Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics, allowing for a fast implementation of C+T. However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
In parallel, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.
Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}. 
{\color{red}
Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}). 
}

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
{\color{red}
In this comparison, we do not include any LMM method, yet, L2-PLR should be very similar to those methods (see section \ref{sec:PLR}).
}
Moreover, we do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For C+T, we report results for a large grid of hyper-parameters. 
For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation. We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data. For example, this underpowered situation occurs when there are many causal variants (3K) to distinguish among many typed variants (280K) while using a small sample size (6K).
}

@@

We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure. This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing''). 
We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary simulation parameters in order to match a range of genetic architectures from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability $h^2$ (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from a model with additive effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $S_\text{causal}$ is the set of causal SNPs, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$ or a Laplace distribution $Laplace(0, \sqrt{h^2 / (2~\vert S_\text{causal} \vert)})$, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).

@@

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000. Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. 
{\color{red}
In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$. This grid-search is directly embedded in our PLR implementation for simplicity. Using $\alpha = 0.001$ should result in a model very similar to gBLUP.\label{sec:PLR} 
}

To fit penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vector of estimated effect sizes is sparse. 
We refer to this method as ``PLR'' in the results section.

@@

We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
When simulating a model with 30 causal SNPs and a heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}). 
Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity and small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for both sexes in the remaining 24K individuals (test set). By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set). 
If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 63.4\% for women and 64.3\% for men.
Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 in the remaining 39K women, while the best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.

@@

The choice of hyper-parameter values is very important since it can greatly impact the performance of methods. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. 
Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
Yet, on the other hand, when choosing a high value for this threshold, too much redundant information is included in the model, which adds some noise to the PRS.
Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated. It means that, in most cases tested in this paper, avoiding redundant information in C+T is more important than including all causal SNPs.
The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T. 
{\color{red}
So, for C+T, we report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
}

In contrast, for the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.

@@

We also explored how to capture non-linear effects. For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects. 
This technique improves the predictive performance of PLR when there are some non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only. 
Moreover, it also improves predictive performance for the celiac disease.

@@

Memory requirements are tightly linked to computation time. Indeed, variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
%There should never be any memory issue thanks to memory-mapping. 
Yet, if too many variables are used in the gradient descent, the OS would regularly swap memory between disk and RAM, severely slowing down computations. 
%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset). Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper. 
Indeed, when using only the 100K most significantly associated SNPs, this reduces correlation between predicted and true heights from 0.657/0.658 to 0.634/0.643 within women/men. 
A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criterions.

@@

Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or ethical reasons (e.g.\ consortium data including many cohorts). As of today, our implementation of penalized logistic regression does not allow for the analysis of summary data, but this represents an important future direction. 
The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.

@@

In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes. A similar penalized linear regression is also available in R package bigstatsr. Our approach solves the dramatic memory and computational burdens faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.

We also demonstrated in simulations and real datasets that our implementation of penalized regressions is highly effective over a broad range of disease architectures. It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease), as well as highly polygenic traits (e.g.\ standing height) providing that sample size is not too small. Finally, PLR as implemented in bigstatsr can also be used to predict phenotypes based on other omics data since our implementation is not specific to genotype data.

@@

Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01). Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
This research has been conducted using the UK Biobank Resource under Application Number 25589.
We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
Finally, we are grateful to the two anonymous reviewers who contributed to improving this paper.

>>>>

\abstract{
Polygenic Risk Scores (PRS) consist in combining genotype information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  
The ``Clumping+Thresholding'' (C+T) approach is the most common method to derive PRS. C+T uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use. 
However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

{\color{red}
Applying penalized linear regression to 350,000 individuals of the UK Biobank, we predict height with a larger correlation than with the best prediction of C+T (\textasciitilde65\% instead of \textasciitilde55\%), further demonstrating its scalability and strong predictive power, even for highly polygenic traits.
Moreover, using 150,000 individuals of the UK Biobank, we are able to predict breast cancer better than C+T, fitting PLR in a few minutes only.
In conclusion, this paper demonstrates the feasibility and relevance of using penalized regression for PRS computation when large individual-level datasets are available, thanks to the efficient implementation available in our R package bigstatsr.
}
}

@@

Polygenic Risk Scores (PRS) consist in combining genotype information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS are useful for genetic epidemiology when testing polygenicity of diseases and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRS. Personalized medicine envisions to use PRS in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
{\color{red}
However, in order to be used in clinical settings, PRS should discriminate well enough between cases and controls. For screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that, for a 10\% disease prevalence, the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.
}

Several methods have been developed to predict disease status, or any phenotype, based on SNP information. A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics, allowing for a fast implementation of C+T. However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
In parallel, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,accuheight2018,mavaddat2019polygenic,okser2014regularized,wei2009disease}.
Finally, Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly polygenic quantitative human phenotypes such as height \cite[]{yang2010common}. 
{\color{red}
Yet, predictions resulting from LMM, known e.g.\ as ``gBLUP'', have not proven as efficient as other methods for predicting several complex diseases based on genotypes (see table 2 of \cite{abraham2013performance}). 
}

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression (PLR), which we compare to the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
{\color{red}
In this comparison, we do not include any LMM method, yet, L2-penalized logistic regression should be very similar to LMM methods (see section \ref{sec:PLR}).
}
Moreover, we do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For C+T, we report results for a large grid of hyper-parameters. 
For PLR, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation. We also use a modified version of PLR in order to capture not only linear effects, but also recessive and dominant effects.

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
{\color{red}
Overall, we find that PLR achieves higher predictive performance than C+T except in highly underpowered cases (AUC values lower than 0.6), while being scalable to biobank data.
}

@@

We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}. This dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure. This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing''). 
We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary simulation parameters in order to match a range of genetic architectures from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability $h^2$ (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from a model with additive effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual as \[y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,\] where $S_\text{causal}$ is the set of causal SNPs, $w_j$ are weights generated from a Gaussian distribution $N(0, h^2 / \vert S_\text{causal} \vert)$ or a Laplace distribution $Laplace(0, \sqrt{h^2 / (2~\vert S_\text{causal} \vert)})$, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), and $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$.
For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).

@@

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium \cite[]{botta2014exploiting}. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000. Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

Finally, for penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function \[L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(z_i\right) + (1 - y_i) \log\left(1 - z_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,\] where $z_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ denotes the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$ is the number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. 
{\color{red}
In this study, we use a grid search over $\alpha \in \{1,~0.5,~0.05,~0.001\}$. This grid-search is directly embedded in our PLR implementation for simplicity. Using $\alpha = 0.001$ should result in a model very similar to gBLUP.\label{sec:PLR} 
}

To fit penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm that does not require to choose this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vector of estimated effect sizes is sparse. 
We refer to this method as ``PLR'' in the results section.

@@

We compared penalized logistic regression (PLR) with the C+T method using simulations of scenario \textnumero1 (Table \ref{tab:simus}).
When simulating a model with 30 causal SNPs and a heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\% for this setting (Figure \ref{fig:main-AUC-logit}). 
Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, except in some cases of high polygenicity and small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-ntrain}).
PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).
For the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

Our implementation of L1-penalized linear regression runs in less than one day for 350K individuals (training set), achieving a correlation of more than 65.5\% with true height for each sex in the remaining 24K individuals (test set). By comparison, the best C+T model achieves a correlation of 55\% for women and 56\% for men (in the test set), and the GWAS part takes one hour (for the training set). 
If using only the top 100,000 SNPs from a GWAS on the training set to fit our L1-PLR, correlation between predicted and true heights drops at 63.4\% for women and 64.3\% for men.
Our L1-penalized logistic regression on breast cancer runs in 13 minutes for 150K women, achieving an AUC of 0.598 in the remaining 39K women, while the best C+T model achieves an AUC of 0.589, and the GWAS part takes 15 hours.

@@

The choice of hyper-parameter values is very important since it can greatly impact the performance of methods. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps. 
For the clumping step, appropriately choosing the $r^2$ threshold is important. 
Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
On the other hand, when choosing a high value for this threshold, too much redundant information is included in the model, which adds noise to the PRS.
Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated. It means that, in most cases tested in this paper, avoiding redundant information in C+T is more important than including all causal SNPs.
The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters for C+T. 
{\color{red}
So, for C+T, we report the best AUC value on the test set, even if it leads to over-optimistic results for C+T as compared to PLR.
}

In contrast, for penalized logistic regression, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$. Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.

@@

We also explored how to capture non-linear effects. For this, we introduced a simple feature engineering technique that enables PLR to detect and learn not only additive effects, but also dominant and recessive effects. 
This technique improves the predictive performance of PLR when there are non-linear effects in the simulations, while providing nearly the same predictive performance when there are additive effects only. 
Moreover, it also improves predictive performance for the celiac disease.

@@

Memory requirements are tightly linked to computation time. Indeed, variables are accessed in memory thanks to memory-mapping when they are used \cite[]{prive2017efficient}. 
When there is not enough memory left, the operating system (OS) frees some memory for new coming variables. 
%There should never be any memory issue thanks to memory-mapping. 
Yet, if too many variables are used in the gradient descent, the OS would regularly swap memory between disk and RAM, severely slowing down computations. 
%If this happens for very large datasets, you can try to use less processing cores (as active sets tend to have non-overlapping variables because of LD) or buy more RAM. 
%Another solution would be to reduce the `n.abort' parameter (default value is 10) in order to stop fitting the regularization path even sooner. 
%This would avoid fitting models with a large number of variables without changing the solution since regularization paths for very large sample sizes are very smooth (Figure \ref{fig:supp-path}).
%We tested processing each chromosome separately and then combining the models later but this did not perform as well as fitting all SNPs at once (data not shown). 
A possible approach to reduce computational burden is to apply penalized regression on a subset of SNPs by prioritizing SNPs using univariate tests (GWAS computed from the same dataset). Yet, this strategy was shown to reduce predictive power \cite[]{abraham2013performance,accuheight2018}, which we also confirm in this paper. 
Indeed, when using only the 100K most significantly associated SNPs, correlation between predicted and true heights is reduced from 0.656/0.657 to 0.634/0.643 within women/men. 
A key advantage of our implementation of PLR is that prior filtering of variables is no more required for computational feasibility, thanks to the use of sequential strong rules and early stopping criteria.

@@

Our approach has one major limitation: the main advantage of the C+T method is its direct applicability to summary statistics, allowing to leverage the largest GWAS results to date, even when individual cohort data cannot be merged because of practical or legal reasons. Our implementation of penalized logistic regression does not allow yet for the analysis of summary data, but this represents an important future direction. 
The current version is of particular interest for the analysis of modern individual-level datasets including hundreds of thousands of individuals.

@@

In this comparative study, we have presented a computationally efficient implementation of penalized logistic regression that can be used to predict disease status based on genotypes. A similar penalized linear regression for quantitative traits is also available in R package bigstatsr. Our approach solves the dramatic memory and computational burdens faced by standard implementations, thus allowing for the analysis of large-scale datasets such as the UK biobank \cite[]{bycroft2017genome}.

We also demonstrated in simulations and real datasets that our implementation of penalized regressions is highly effective over a broad range of disease architectures. It can be appropriate for predicting autoimmune diseases with a few strong effects (e.g.\ celiac disease), as well as highly polygenic traits (e.g.\ standing height) provided that sample size is not too small. Finally, PLR as implemented in bigstatsr can also be used to predict phenotypes based on other omics data since our implementation is not specific to genotype data.

@@

Authors acknowledge LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01) and ANR project FROGH (ANR-16-CE12-0033). Authors also acknowledge the Grenoble Alpes Data Institute that is supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02).
This research has been conducted using the UK Biobank Resource under Application Number 25589.
We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.
Finally, we are grateful to the two anonymous reviewers who contributed to improving this paper.
